Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-07-01 workflow/workflow.Rmd
compare ours and published

- solve loading packages issues (DR),
- versions of packages we want to use in develop branches and change functions in text accordingly (EP and DR),
- clean up singlecellworkflow package (FP),
- update NAMESPACE and DESCRIPTION (FP),
- install and run package using cache instead of saved files for zinbwave and clusterExperiment and making sure we can build the vignette using R 3.4 and bioc-devel 3.6 (all?),
- make singlecellworkflow package public (FP).]

@@

# Create metadata data.frame
metadata <- data.frame("Experiment" = bio,
                       "Batch" = batch,
                       "clusterLabels" = clusterLabels[m,2],
                       qc)

# Symbol for cells not assigned to a lineage in original data
metadata$clusterLabels[is.na(metadata$clusterLabels)] <- -2

@@

```{r}
load("../data/oe_se_1000Var.rda")
```

@@

```{r original}
clus.labels <- colData(core)[, "clusterLabels"]
col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
col_clus <- col_clus[1:length(unique(clus.labels))]
names(col_clus) <- sort(unique(clus.labels))
table(clus.labels)
```

@@

```{r clustbatch}
table(data.frame(batch = as.vector(batch),
                 cluster = clus.labels))
```

@@

```{r pcanorm, fig.cap="PCA of normalized expression measures, where each point represents a cell. In the left panel, cells are color-coded by batch. In the right panel, cells are color-coded by original published clusters."}
pca <- prcomp(t(norm))
par(mfrow = c(1,2))
plot(pca$x, col = col_batch[batch], pch = 20, main = '')
plot(pca$x, col = col_clus[as.character(clus.labels)], pch = 20, main = '')
```

@@

```{r mdsW, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clusters."}
W <- colData(se)[, grepl('^W', colnames(colData(se)))]
W <- as.matrix(W)
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(clus.labels)], main = '',
     pch = 20, xlab = 'Component 1', ylab = 'Component 2')
legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
       fill = unique(col_clus), title = 'Sample')
```

@@

```{r rsec_50}
fn <- '../data/RSEC_W_combineMinSize10.rda'
if (runClus & !file.exists(fn)){
  # Symbol for samples missing from original clustering
  seObj <- SummarizedExperiment(t(W), colData = colData(core))
  print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1),
                                  betas = 0.8, dimReduce="none",
                clusterFunction = "hierarchical01", minSizes=1,
                ncores = NCORES, isCount=FALSE,
	       dendroReduce="none",dendroNDims=NA,
                subsampleArgs = list(resamp.num=100,
                                     clusterFunction="kmeans",
                                     clusterArgs=list(nstart=10)),
                verbose=TRUE,
                combineProportion = 0.7,
                mergeMethod = "none", random.seed=424242,
                combineMinSize = 10)))
  save(ceObj, file = fn)
}else{
  load(fn)
}
```

@@

The distribution of cells across the consensus clusters is as follows:

@@

@@

 ```{r addPublishedClusters}
 #delete this chunk if not of interest
 #if keep, need to choose if do both figures and add figure caption 
 ceObj<-addClusters(ceObj,colData(ceObj)$clusterLabels,clusterLabel="publishedClusters")
 plotClusters(ceObj, whichClusters=c("mergeClusters","combineMany","publishedClusters"))
 plotBarplot(ceObj,whichClusters=c("mergeClusters","publishedClusters")) 
 ```

 EP: If we do the above code, then the `plotHeatmap` command below should be modified to have the `publishedClusters` under the `whichClusters` argument.

@@

```{r heatmapsClusters, fig.cap="Heatmap of the normalized expression measures for the 1,000 most variable genes, where rows correspond to genes and columns to cells ordered by RSEC clusters."}
# Set colors for cell clusterings
colData(ceObj)$clusterLabels <- as.factor(colData(ceObj)$clusterLabels)
origClusterColors <- bigPalette[1:nlevels(colData(ceObj)$clusterLabels)]
experimentColors <- bigPalette[1:nlevels(colData(ceObj)$Experiment)]
batchColors <- bigPalette[1:nlevels(colData(ceObj)$Batch)]
metaColors <- list("Experiment" = experimentColors,
                   "Batch" = batchColors,
                   "clusterLabels" = origClusterColors)

plotHeatmap(ceObj, visualizeData = assays(se)$normalizedValues,
            whichClusters = "primary", clusterFeaturesData = "all",
            clusterSamplesData = "dendrogramValue", breaks = 0.99,
            sampleData = c("clusterLabels", "Batch", "Experiment"),
            clusterLegend = metaColors, annLegend = FALSE, main = '')
```

@@

```{r tabagain}
table(data.frame(original = clus.labels, ours = primaryClusterNamed(ceObj)))
```

>>>>

- solve loading packages issues (DR),
- versions of packages we want to use in develop branches and change functions in text accordingly (EP and DR),
- update clustering section with new clusterExperiment functions (EP),
- add details in section preprocessing (DR),
- address Sandrine's comments in sections slingshot and de (KS),
- clean up singlecellworkflow package (FP),
- update NAMESPACE and DESCRIPTION (FP),
- install and run package using cache instead of saved files for zinbwave and clusterExperiment and making sure we can build the vignette using R 3.4 and bioc-devel 3.6 (all?),
- make singlecellworkflow package public (FP).]

@@

# Create metadata data.frame
metadata <- data.frame("Experiment" = bio,
                       "Batch" = batch,
                       "publishedClusters" = clusterLabels[m,2],
                       qc)

# Symbol for cells not assigned to a lineage in original data
metadata$publishedClusters[is.na(metadata$publishedClusters)] <- -2

@@

```{r}
load("../data/oe_se_1000Var.rda")
colnames(colData(core))[colnames(colData(core)) == 'clusterLabels'] = 'publishedClusters' 
```
[FP: delete this chunk once we can load scone]

@@

```{r original}
publishedClusters <- colData(core)[, "publishedClusters"]
col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
col_clus <- col_clus[1:length(unique(publishedClusters))]
names(col_clus) <- sort(unique(publishedClusters))
table(publishedClusters)
```

@@

```{r clustbatch}
table(data.frame(batch = as.vector(batch),
                 cluster = publishedClusters))
```

@@

```{r pcanorm, fig.cap="PCA of normalized expression measures, where each point represents a cell. In the left panel, cells are color-coded by batch. In the right panel, cells are color-coded by original published clusters."}
pca <- prcomp(t(norm))
par(mfrow = c(1,2))
plot(pca$x, col = col_batch[batch], pch = 20, main = '')
plot(pca$x, col = col_clus[as.character(publishedClusters)], pch = 20, main = '')
```

@@

```{r mdsW, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clusters."}
W <- colData(se)[, grepl('^W', colnames(colData(se)))]
W <- as.matrix(W)
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(publishedClusters)], main = '',
     pch = 20, xlab = 'Component 1', ylab = 'Component 2')
legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
       fill = unique(col_clus), title = 'Sample')
```

@@

```{r rsec_50}
fn <- '../data/RSEC_W_combineMinSize10.rda'
if (runClus & !file.exists(fn)){
  # Symbol for samples missing from original clustering
  seObj <- SummarizedExperiment(t(W), colData = colData(core))
  print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1),
                                  betas = 0.8, dimReduce="none",
                clusterFunction = "hierarchical01", minSizes=1,
                ncores = NCORES, isCount=FALSE,
	       dendroReduce="none",dendroNDims=NA,
                subsampleArgs = list(resamp.num=100,
                                     clusterFunction="kmeans",
                                     clusterArgs=list(nstart=10)),
                verbose=TRUE,
                combineProportion = 0.7,
                mergeMethod = "none", random.seed=424242,
                combineMinSize = 10)))
  save(ceObj, file = fn)
}else{
  load(fn)
  colnames(colData(ceObj))[colnames(colData(ceObj)) == 'clusterLabels'] = 'publishedClusters'
}
```

@@

The distribution of cells across the consensus clusters can be visualized in Figure \@ref(fig:barplotOurs)) and is as follows:

@@

```{r barplotOurs, fig.cap="Barplot of number of cells per cluster found using our workflow."}
plotBarplot(ceObj) 
```

[FP: Should we add here the short paragraph Elizabeth was talking about to compare our found clusters versus the original published clusters?]

```{r addPublishedClusters, fig.cap="Barplot of number of cells per cluster in original published clusters."}
ceObj<-addClusters(ceObj, colData(ceObj)$publishedClusters, 
                   clusterLabel = "publishedClusters")
plotBarplot(ceObj,whichClusters=c("mergeClusters","publishedClusters")) 
```

@@

[FP: after the modifications, do we still need function addCluters? If we want to illustrate that clusterExperiment package has a convenient function to add clusters to a clusterExperiment object, we can keep it and say that here it is not necessary.]

```{r comparePublished, fig.cap="Comparison between the clusters found using our workflow and original published clusters."}
plotClusters(ceObj, whichClusters=c("mergeClusters","combineMany","publishedClusters"))
```

[EP: If we do the above code, then the `plotHeatmap` command below should be modified to have the `publishedClusters` under the `whichClusters` argument. ]

[FP: Elizabeth, I changed the `clusterLabels` to `publishedClusters`, more informative indeed. Maybe, we could add the plotClusters to compare the published and our own clusters and the plotBarplot to illustrate our previous call to function table()? I modified the code accordingly. If you don't like it, feel free to change, it is just a proposition.]

@@

```{r heatmapsClusters, fig.cap="Heatmap of the normalized expression measures for the 1,000 most variable genes, where rows correspond to genes and columns to cells ordered by RSEC clusters."}
# Set colors for cell clusterings
colData(ceObj)$publishedClusters <- as.factor(colData(ceObj)$publishedClusters)
origClusterColors <- bigPalette[1:nlevels(colData(ceObj)$publishedClusters)]
experimentColors <- bigPalette[1:nlevels(colData(ceObj)$Experiment)]
batchColors <- bigPalette[1:nlevels(colData(ceObj)$Batch)]
metaColors <- list("Experiment" = experimentColors,
                   "Batch" = batchColors,
                   "publishedClusters" = origClusterColors)

plotHeatmap(ceObj, visualizeData = assays(se)$normalizedValues,
            whichClusters = "primary", clusterFeaturesData = "all",
            clusterSamplesData = "dendrogramValue", breaks = 0.99,
            sampleData = c("publishedClusters", "Batch", "Experiment"),
            clusterLegend = metaColors, annLegend = FALSE, main = '')
```

@@

```{r tabagain}
table(data.frame(original = publishedClusters, ours = primaryClusterNamed(ceObj)))
```
